---
title: 'Evaluation of Fine Mapping methods using the R package qgg'
author: "Merina Shrestha, Palle Duun Rohde & Peter Sørensen"
date: "`r Sys.Date()`"
bibliography: qg2021.bib
biblio-style: apalike
link-citations: yes
output:
  pdf_document:
    dev: png
    includes:
      in_header: preamble.tex
  html_document:
    includes:
      in_header: mathjax_header.html
---


```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE, tidy.opts = list(width.cutoff = 70), tidy = TRUE)
eval <- FALSE
echo_solution <- FALSE
```

# Introduction

Comparing finemapping methods using Bayesian Linear regression models


#### Software used: 
To follow the practical, you will need the following installed (see installation guides below):

* R (version $\geq$ 4.2)
* qgg (version $\geq$ 1.1.1)


### Input data/objects used for finemapping in the `qgg` package
All functions in `qgg` used for finemapping based on BLR models relies on a simple data infrastructure that takes five main input: 

`y`:\qquad \quad vector, matrix or list of phenotypes\newline
`X`:\qquad \quad design matrix for non-genetic factors \newline
`W`:\qquad \quad matrix of centered and scaled genotypes (in memory) \newline
`Glist`:\quad list structure providing information on genotypes, sparse LD, and LD scores (on disk) \newline
`stat`:\quad data frame with marker summary statistics \newline


### Load necessary packages
```{r,  eval=TRUE, echo=TRUE}
library(qgg)
```


### Prepare genotype information in Glist format
\vspace{-1truemm}
```{r,  eval=TRUE, echo=TRUE}
bedfiles <- system.file("extdata", paste0("sample_chr",1:2,".bed"), package = "qgg")
bimfiles <- system.file("extdata", paste0("sample_chr",1:2,".bim"), package = "qgg")
famfiles <- system.file("extdata", paste0("sample_chr",1:2,".fam"), package = "qgg")

Glist <- gprep(study="1000G", bedfiles=bedfiles, bimfiles=bimfiles, famfiles=famfiles)

```

### Filter markers
\vspace{-1truemm}
```{r,  eval=TRUE, echo=TRUE}
rsids <-  gfilter(Glist = Glist, excludeMAF=0.05, excludeMISS=0.05, excludeHWE=1e-12) 
```

### Compute sparse LD matrices and ldscores
\vspace{-1truemm}
```{r,  eval=TRUE, echo=TRUE}
ldfiles <- paste0("C:\\Users\\au223366\\Dropbox\\Projects\\qgg2.0\\1000G\\chr",1:2,".ld")
Glist <- gprep(Glist, task="sparseld", msize=200, rsids=rsids, ldfiles=ldfiles, overwrite=TRUE)
```

### Simulate data
\vspace{-1truemm}
```{r,  eval=TRUE, echo=TRUE}
sim <- gsim(Glist=Glist,nt=1)
W <- getG(Glist=Glist, chr=1, rsids=Glist$rsidsLD[[1]], scale=TRUE)
sim <- gsim(W=W,nt=1)

```


### Compute summary statistics (e.g. fit single marker regression model)
\vspace{-1truemm}
```{r,  eval=TRUE, echo=TRUE}
stat <- glma(y=sim$y, Glist=Glist)  
```


### BLR model analysis based on summary statistics
\vspace{-1truemm}
```{r,  eval=TRUE, echo=TRUE}
fit <- gbayes( stat=stat, Glist=Glist, method="bayesC", nit=1000)
plot(fit[[1]]$dm)
```

### Mapping based on stat defined by variance per marker
\vspace{-1truemm}
```{r,  eval=TRUE, echo=TRUE}
# extract marker variance
stat <- (fit$stat$bm^2)*fit$stat$af*(1-fit$stat$af)*2
names(stat) <- rownames(fit$stat)

# define sets
msize <- 20
sets <- qgg:::splitWithOverlap(names(stat), msize, msize-1)

# gsea using sum
nperm <- 1000
mma <- gsea(stat=stat,sets=sets, nperm=nperm)
mma[mma[,"p"]==0,"p"] <- 0.1/nperm
head(mma)
plot(mma[,"stat"],-log10(mma[,"p"]))
```

### Mapping based on stat defined by posterior inclusion probability
\vspace{-1truemm}
```{r,  eval=TRUE, echo=TRUE}
# extract pip
stat <- fit[[1]]$dm
names(stat) <- names(fit[[1]]$bm)

# define sets
msize <- 20
sets <- qgg:::splitWithOverlap(names(stat), msize, msize-1)

# gsea using sum
nperm <- 1000
mma <- gsea(stat=stat,sets=sets, nperm=nperm)
mma[mma[,"p"]==0,"p"] <- 1/nperm
head(mma)
plot(-log10(mma[,"p"]))
plot(mma[,"stat"],-log10(mma[,"p"]))
```

### Mapping based on credible sets
Bayesian methods can be used to determine the credible set, the minimum set 
of SNPs that contains all causal SNPs with probability alpha. 
When assuming only one causal SNP, alpha is the sum of PIPs for SNPs in a set. 
This means that an alpha credible set is equivalent to ranking SNPs from largest to smallest 
PIPs and taking the cumulative sum of PIPs until it is at least alpha (https://doi.org/10.1038/s41576-018-0016-z).

\vspace{-1truemm}
```{r,  eval=TRUE, echo=TRUE}
# extract pip
stat <- fit[[1]]$dm
names(stat) <- names(fit[[1]]$bm)

# define sets
msize <- 20
sets <- qgg:::splitWithOverlap(names(stat), msize, msize-1)

# compute credible sets 
alpha <- 0.9
cs <-sapply(sets,function(x){
  x <- x[x%in%names(stat)]
  x <- stat[x]
  x <- sort(x, decreasing=TRUE)
  x <- cumsum(x)
  sum(x>alpha)
})
str(cs)
plot(cs)
```

